Synthesizing Quotient Lenses

### 2018-03-14 papers/QuotientLenses/main.tex
minor nit fixes in section 4

\section{Quotient Regular Expressions}
\label{QRE}
A Quotient Regular Expression (or QRE) is a regular expression $R$ augmented
with syntax that expresses an equivalence relation on the language of $R$. This
section formalizes the set of QRE combinators that we introduced informally
in \cref{sec:example}.
\iflastminute 
\saz{Why does the ``cref'' macro not capitalize Section?}
\bcp{Dunno.  Because SIGPLAN does not want it to??  But maybe we should be
  using autoref instead anyway...}
\fi

@@

\begin{tabular}{rl}
  $W(Q)$       & A regular expression, denoting the ``whole'' of $Q$\\
  $\eqrel{Q}$  & An equivalence relation on $\mathcal{L}(W(Q))$\\
  $K(Q)$       & A regular expression, denoting the ``kernel'' of $Q$, such that
                  $\mathcal{L}(K(Q))$ forms a \\
               & complete set of representatives for $\eqrel{Q}$ \\
  $\canonize(Q)$ & 
  A ``canonizing'' function. Given any $w \in \mathcal{L}(W(Q))$,
  $\canonize(Q)(w)$ \\ & is the unique $k$ in $\mathcal{L}(K(Q))$ such
  that $k$ is equivalent to $w$ mod $\eqrel{Q}$
\end{tabular}
  % of type $\mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$  which

@@

Semantically, the \kw{normalize} QRE is universal---each of the other
combinators $Q$ is equivalent to $\normalize{W(Q)}{K(Q)}{f}$ for some
surjective, idempotent function 
$f : \mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$.  
However,
verifying that a canonization
function $f$ is surjective and idempotent is in general undecidable.
Consequently, a programmer wishing to use \kw{normalize} must
discharge strong proof obligations, which is cumbersome in
practice.\footnote{In our implementation we allow a programmer to
use \kw{normalize} at their own risk without checking these side
conditions as an ``escape hatch'' for the case when other QRE
combinators are insufficient.\bcp{Do we use this in any of the
programs we've written?}}

@@

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonize{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonize(\id(R)) &=& id_{\mathcal{L}(R)} \\
\canonize(\collapse{R}{s})(w) &=& s\\[1ex]
\canonize(\squash{Q_1}{Q_2}{f})(w) &=&
\begin{cases}
\canonize(Q_1)(f(w)) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{otherwise}
\end{cases}\\[3ex]
\canonize(\normalize{R_1}{R_2}{f}) &=& f\\
\canonize(Q_1 \; ; \; Q_2) &=& \canonize(Q_2) \circ \canonize(Q_1)\\
\canonize(Q_1 \cdot Q_2) &=& \canonize(Q_1) \cdot \canonize(Q_2)\\[1ex]
\canonize(Q_1 \sep Q_2)(w) &=&
\begin{cases}
\canonize(Q_1)(w) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{if } w \in \mathcal{L}(W(Q_2))\\
\end{cases}\\[3ex]
\canonize(Q^*) &=& \canonize(Q)^* \\
\end{array}
\]
\end{center}
\begin{align*}
&\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})\\
&= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q)(s_{n-1}))\cdot
(\canonize(Q_n)(r_n))
\end{align*}
\caption{QRE canonizers. \ksf{Solomon: please define/explain the . and *
operators on functions.}}
\label{fig:canonizers}
\end{figure}
The \kw{permutation} combinator gives rise to the most interesting
definition:
\[
\begin{array}{cl}
 & \canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= & (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q)(s_{n-1}))\cdot
(\canonize(Q_n)(r_n))
\end{array}
\]
\noindent 
which places the strings that match $Q_i$ and $Q$ according to the
canonical permutation (i.e., the identity permutation) before applying
the canonizers $\canonize(Q_i)$ and $\canonize(Q)$, respectively.

@@

Strong unambiguity is necessary in the premises of the regular QRE combinators
because it ensures that the canonizing function of a QRE is well-defined.
For example, consider the QRE \lstinline{"a"* . (collapse "a"* $\to$ "a")}. The
behaviour of this QRE is not well-defined since the string
\cd{"aaa"} can be canonized to any of \cd{"a"}, \cd{"aa"}, or \cd{"aaa"}
depending on how \cd{"aaa"} is parsed. We also require that the regular
combinators applied to the kernels of QREs be unambiguous since the underlying
bijective string lens of a QRE lens operates on kernels, and bijective string lenses
impose the same unambiguity restrictions so that they too are well defined as
functions.

>>>>

\section{Quotient Regular Expressions}
\label{QRE}
A Quotient Regular Expression (or QRE) is a regular expression $R$ augmented
with syntax that expresses an equivalence relation on the language of $R$. This
section formalizes the set of QRE combinators that we introduced informally
in \cref{sec:example}.

@@

\begin{tabular}{rl}
  $W(Q)$       & A regular expression, denoting the ``whole'' of $Q$\\
  $\eqrel{Q}$  & An equivalence relation on $\mathcal{L}(W(Q))$\\
  $K(Q)$       & A regular expression, denoting the ``kernel'' of $Q$, such that
                  $\mathcal{L}(K(Q))$ forms a \\
               & complete set of representatives for $\eqrel{Q}$ \\
  $\canonize(Q)$ & 
  A ``canonizing'' function. Given any $w \in \mathcal{L}(W(Q))$,
  $\canonize(Q)(w)$ \\ & is the unique $k$ in $\mathcal{L}(K(Q))$ such
  that $k \eqrel{Q} w$.
\end{tabular}
  % of type $\mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$  which

@@

Semantically, the \kw{normalize} QRE is universal---each of the other
combinators $Q$ is equivalent to $\normalize{W(Q)}{K(Q)}{f}$ for some
surjective, idempotent function 
$f : \mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$.  
However,
verifying that a canonization
function $f$ is surjective and idempotent is in general undecidable.
Consequently, a programmer wishing to use \kw{normalize} must
discharge strong proof obligations, which is cumbersome in
practice.\footnote{In our implementation we allow a programmer to
use \kw{normalize} at their own risk without checking these side
conditions as an ``escape hatch'' for the case when other QRE
combinators are insufficient.\bcp{Do we use this in any of the
programs we've written?}\afm{Nope.}}

@@

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonize{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonize(\id(R)) &=& id_{\mathcal{L}(R)} \\
\canonize(\collapse{R}{s})(w) &=& s\\[1ex]
\canonize(\squash{Q_1}{Q_2}{f})(w) &=&
\begin{cases}
\canonize(Q_1)(f(w)) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{otherwise}
\end{cases}\\[3ex]
\canonize(\normalize{R_1}{R_2}{f}) &=& f\\
\canonize(Q_1 \; ; \; Q_2) &=& \canonize(Q_2) \circ \canonize(Q_1)\\
\canonize(Q_1 \cdot Q_2)(w_1 \cdot w_2) &=& \canonize(Q_1)(w_1) \cdot \canonize(Q_2)(w_2)\\[1ex]
\canonize(Q_1 \sep Q_2)(w) &=&
\begin{cases}
\canonize(Q_1)(w) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{if } w \in \mathcal{L}(W(Q_2))\\
\end{cases}\\[3ex]
\canonize(Q^*)(\epsilon) &=& \epsilon \\
\canonize(Q^*)(w_1 \cdot w_2) &=& \canonize(Q)(w_1)\cdot\canonize(Q^*)(w_2) \\
\end{array}
\]
\end{center}
\begin{align*}
&\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})\\
&= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q)(s_{n-1}))\cdot
(\canonize(Q_n)(r_n))
\end{align*}
\caption{QRE canonizers. \ksf{Solomon: please define/explain the . and *
operators on functions.}\afm{I tried rewriting with explicit arguments,
thoughts?}}
\label{fig:canonizers}
\end{figure}
The \kw{permutation} combinator gives rise to the most interesting
definition:
\[
\begin{array}{cl}
 & \canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= & (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q)(s_{n-1}))\cdot
(\canonize(Q_n)(r_n))
\end{array}
\]
\noindent 
which places the strings that match $Q_i$ and $Q$ according to the
canonical permutation (i.e., the identity permutation) before applying
the canonizers $\canonize(Q_i)$ and $\canonize(Q)$, respectively.

@@

Strong unambiguity is necessary in the premises of the regular QRE combinators
because it ensures that the canonizing function of a QRE is well-defined.
For example, consider the QRE \lstinline{"a"* . (collapse "a"$^*$ $\to$ "a")}. The
behaviour of this QRE is not well-defined since the string
\cd{"aaa"} can be canonized to any of \cd{"a"}, \cd{"aa"}, or \cd{"aaa"}
depending on how \cd{"aaa"} is parsed. We also require that the regular
combinators applied to the kernels of QREs be unambiguous since the underlying
bijective string lens of a QRE lens operates on kernels, and bijective string lenses
impose the same unambiguity restrictions so that they too are well defined as
functions.
